---
title: "Project 1"
author: "Alice Friedman"
date: "9/23/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
library(dplyr)
library(tidyr)
```
# Step 1: Load required libraries and read text file into R
Code that follows requires libraries stringr, dplyr, and tidyr.
```{r}
library(stringr)
library(dplyr)
library(tidyr)

raw.data <- read.delim("https://raw.githubusercontent.com/aliceafriedman/DATA607_Proj1/master/tournamentinfo.txt", header = FALSE, sep = "|", dec = ".")
```
##Step 2: Read data from each series of rows into its own table
The data is spread into 2 rows. The following code splits the data into the first row and second row for variables pertaining to each player.
```{r}
head(raw.data)

#code checks for the first row by looking for the player's ID, which is an integer
first_row_test <- str_detect(raw.data$V1, "[[:digit:]]{1,3}")
#code subsets rows meeting the first_row_test, starting with the first row of player data 
first_rows <- raw.data[first_row_test,]
#code checks for the second row by matching a 2 letter code regex to find the rows with state info
second_row_test <- str_detect(raw.data$V1, "[[:upper:]]{2}")
#code subsets rows meeting the second_row_test, starting with the first row of player data
second_rows <- raw.data[second_row_test,]

head(first_rows)
tail(first_rows)
head(second_rows)
tail(second_rows)
```
  
##Step 3: Combine data into single dataframe and add coloumn names.
Then the data is recombined into a single dataframe.

```{r}
#extract column names
colnames1 <- raw.data[2,]
colnames2 <- raw.data[3,]
newcolnames <- cbind(colnames1, colnames2)

#This removes extra spaces before and after the names
newcolnames <- lapply(newcolnames, trimws)

#create dataframe where each player's data in on one row
df <- cbind(first_rows, second_rows)

#convert column names to character
#note: this step is important or colnames function wil read the factor indices insted of the text
newcolnames <- lapply(newcolnames, as.character)

#apply original colnames--now on a single row--to dataframe
colnames(df) = newcolnames

#view results
head(df)

#This results in too many columns named "Round!" Need to rename
new_points_names <- paste0("Round ", colnames(df[,15:21]), " Pair")
colnames(df)[4:10] = new_points_names
new_round_names <- paste0("Round ", c(1:7))
colnames(df)[15:21] = new_round_names

#Drop empty columns
df <- df[c(1:10,12:21)]

#This produces a dataframe with approximately the original column names
head(df)
names(df)
```
  
##Step 4: Calculate opponents average pre-tournament scores.
```{r}
#Create a subset of the data frame that contains all the info that can be extracted without a join
df_sub <- select(df, Pair, "Player Name", Num, Total, "USCF ID / Rtg (Pre->Post)") %>%  
  rename("PairID"=Pair, "Player Name"="Player Name", "State"=Num, "Total Points"=Total, "ID_Ratings"="USCF ID / Rtg (Pre->Post)") %>%  
#Extract Pre- ratings from combined ID_Ratings column, based on #StartEnd <- str_locate(df_sub$ID_Ratings, "R:( )+?[[:digit:]]{3,4}"
  mutate(Pre.Rating = str_sub(ID_Ratings, start=15)) %>%
  mutate_all(str_trim) %>%
  mutate(Pre.Rating = as.numeric(str_extract(Pre.Rating,"[[:digit:]]+"))) %>%
  arrange(Pre.Rating) %>%
#Drop ID_Ratings
  select(-ID_Ratings) %>%
  glimpse()
```

```{r}
#Reshape data into long format using tidyr
#Create new long-format table with pair information
df_pairs <- df %>% select(ends_with("Pair")) %>%
  gather(key = "Round", value = "PairID", -Pair, na.rm = FALSE, convert = FALSE, factor_key = TRUE) %>%
  arrange(Pair) %>%
  mutate(PairID = str_extract(PairID, "[[:digit:]]+")) %>%
  glimpse()

#Join with df_sub to get the Player.2's ratings
df_player2_avg_rating <- inner_join(df_pairs, df_sub) %>%
#Group by pair
  group_by(Pair) 

#Summarize to  obtain mean rating of opponents
Player.2.Mean <-  df_player2_avg_rating %>% summarize("Average Opponent Rating" = as.integer(mean(Pre.Rating))) %>%
  mutate (PairID = as.character(Pair)) %>%
  mutate_all(str_trim) %>%
  mutate_all(unlist) %>%
  glimpse()
```
   
##Final Step: Rejoin to dataframe for final answer, and write to CSV!
```{r}
#Rejoin to df_sub for final answer!
Chess <- inner_join(df_sub, Player.2.Mean) %>%
  select(-Pair) %>%
  arrange(PairID) %>%
  mutate(`Total Points`=as.numeric(`Total Points`), 
         Pre.Rating = as.numeric(Pre.Rating),
         `Average Opponent Rating`=as.numeric(`Average Opponent Rating`)) %>%
  glimpse() 

head(Chess)

#Note file will write to location set up in your RStudio preferences.
write.csv(Chess, file="ChessScores.csv")
```
  
##Data Visualization
Requires ggplot2
```{r}
library(ggplot2)

#What is the relationship between Total Points (e.g. how the player did at this tournament) and Pre-Tournament Rating?
Chess %>% 
  ggplot(aes(x=Pre.Rating, y=`Total Points`))+
  geom_point()+
  geom_smooth(method = "lm")+
  labs(title="Pre-Tournament Ratings vs. Tournament Total Points",
       x="Pre-Tournament Ratings",
       y="Tournament Total Points")


#What is the relationship between Total Points (e.g. how the player did at this tournament) and Average Opponent Pre-Tournament Rating?
Chess %>% 
  ggplot(aes(x=`Average Opponent Rating`, y=`Total Points`))+
  geom_point()+
  geom_smooth(method = "lm")+
  labs(title="Mean Opponent Ratings vs. Tournament Total Points",
       x="Mean Oppoenent Pre-Tournament Ratings",
       y="Tournament Total Points")
```
  
Not surpringly, better-rated players did better at this tournament. Somewhat surprisingly, players paired with better-rated opponents *also* did better at this tournament! This can likely be explained that as players advanced, they were more likely to be paired with their better-rated peers, who had also won the last round.

